\section{Fear}

Every time I introduce myself at a party, my new acquaintance asks,
``So, what do you do?''  ``I'm a physicist. I teach physics and math
at a university,'' I say, and then I pause, because just about every
time, the reaction is the same: ``Oooo --- you must be very smart.  I
was \emph{terrible} at that.''  And since I'm a slightly awkward geek,
I stammer out my assurances that I'm not that smart, and then I
commiserate about how difficult the high school/college versions of
science or math classes usually are. 


I do a bit better with my students, where I have more time to talk,
but the results are frequently the same: A large percentage of the
people I meet claim to have liked math and science at some point in
their lives, they even did well at it once upon a time.  But then
something happened.  When that something occurred shows a lot of
variation: in middle school, in high school, in college.  But at some
point, most people decide that science and math are not for them --
they feel that they have been \emph{told}, rudely, that they are not
\emph{good enough} for math and science.  They want to work hard in my
class, they assure me, but I have to understand that this stuff just
doesn't come naturally to them.


So I start every term with a problem, as soon as the students walk in
the room.  It goes something like this:

\subsection{Goats and Sports Cars}
Note: This is a famous problem in mathematics.  If you've seen it
before and know the answer, keep it to yourself for now.  If you
haven't seen it before or you don't know the answer, DON'T LOOK IT UP!
The point isn't the answer, it is the process of getting the answer.

Suppose you are on a game show, and the host offers you an opportunity
to select a prize from behind one of three doors.  The prize (a sports
car) is behind one of them.  There are goats behind the other two.  So
you select one of the doors.

Now stop right there for a moment, and think about the choice you just
made.  What is the probability that you were right on your first
guess?  Don't click on the ``Answer'' link below until you are
satisfied that you have an answer.  You can use the computation box
below to do some figuring, if you have to.


\begin{sageverbatim}
#You can use this cell like a calculator.  
#Type in a calculation, and hit ``Shift-enter'' to evaluate.
\end{sageverbatim}


%right here I want a ``hidden'' sage cell, where the hidden tag is
%replaced with ``answer''.
%TODO: Change the display of the word ``hide'' to ``answer''.
\begin{sageverbatim}
%hide
%latex
#The answer is $\frac{1}{3}$, since there are three choices, and one has the prize.
\end{sageverbatim}

\begin{sageverbatim}
%hide
%auto
@interact
def _(Probability=([0/3,1/3, 2/3, 3/3])):
    choice = Probability
    if choice == 1/3:
         html('1 / 3 is right.')
    else:
         html('%s is not right.'%(choice))
\end{sageverbatim}    





But now, like any good game show host, I'm going to throw you a
curveball: I don't want the game to end so quickly.  There's not
enough time run more commercials, or to play that awful music, unless
I make you make you sweat some more. 


So here's the play: I announce that I know where the prize is, and I'm
going to help you out by showing you that the prize is NOT behind one
of the doors you didn't choose (I wouldn't expose the one you chose,
because then the game would be over.  No Fun.  And I won't expose one
where the prize IS, because, again, No Fun. And since we know that
there's only one prize, I can always find a door to open that won't
ruin the game.  Let's say you guessed Door Number 1, and then I open
Door Number 3, and show you that there's a goat back there.  And then
I ask, with as much cheesiness as I can muster: ``Would you like to
switch?  To Door Number 2?  Or would you like to stick with your
original guess of Door Number 1?''


What do you do?  Is it better to switch, or to stay?  Or does it even
matter?


I recommend trying it out.  Write down your best guess as to the
correct strategy (the strategy that leads you to win the sports car
most often) and an explanation of WHY that strategy is the best one.
Then play the game.  Pretend to be the host, have your roommate be the
contestant, and keep track of how many times you win if you stay with
your original guess, and how many times you win if you switch.


\subsection{The answer}
Alright, so I assume you tried it out, and since this is a game that
involves some randomness, I hope you tried it a bunch of times,
because we all know that it is possible to flip a coin 10 times in a
row Heads -- it is only when you try it hundreds of times that you can
be sure the pattern of half and half will be observed.

What did you find?

Almost everyone makes an initial answer that it doesn't matter which
door you pick.  The reasoning is incredibly compelling: There are two
doors.  We have no reason to favor one over the other, therefore the
prize has an equal probability to be behind either door, therefore
there is no advantage in switching or in staying.

Incredibly compelling, but wrong.

Once you do the experiment by playing the game enough times, almost
everyone becomes convinced that in fact there is a $\frac{2}{3}$rds
chance of winning if you switch, and only $\frac{1}{3}$rd if you stay,
because you win more than 60\% of the time if you switch, and only
around 30\% of the time if you stay.

And here's the interesting thing: most people still have a hard time
putting into words why the answer SHOULD BE that it is better to
switch.  They can accept, empirically, that it seems to come out that
way, but lacking a good explanation of why, many people come away from
this problem confused.

I personally took a couple of weeks to understand this problem, when I
first saw it in Discrete Mathematics as an undergraduate.  The reason
the problem is famous is that it was discussed in a popular magazine
article in the early 1990's, and dozens of mathematics professors
wrote into the author, claiming that she had gotten it wrong when in
fact she was right.  Most of them wrote sincere apologies later.

Almost everyone gets this problem wrong at first, and even after
seeing a good explanation, many of them still have a hard time
accepting it.

I'm not going to go into an explanation of the correct answer here,
because I've found that it mostly doesn't help people ot \emph{hear}
(or read) the answer.  It helps them to DO it.  And then think through
it, and make sense of the problem themselves.

Instead, what I want to do is talk about what this problem, and
people's reactions to it, tells us about Mathematics.

\begin{itemize}
\item \textbf{When you do math, you will be WRONG most of the time.}

Most people have a really hard time with being told they are wrong.
And somehow math teachers are, let's say, \emph{gifted} at making
other people feel stupid.  It is not intentional, I assure you.  I am
a math teacher, and a lot of my friends are math teachers.  And as a
group, we are usually well-meaning, friendly, thoughtful people who
really want to help you succeed.  But somehow we come across as the
most condescending jerks in the academic world.

I think this happens because of Fact 1.  It is very easy to tell when
something is wrong in math.  The numbers and equations provide a handy
truth-check that makes nonsense just leap off the page, once you know
what to look for.  Math teachers see that kind of thing every single
day, so we are very good at spotting errors.  And pointing them out.
And that makes people feel dumb.  And no one likes to feel dumb.
Therefore, lots of people don't like math.

But being wrong most of the time is actually one of the most important
and useful features of math.  One doesn't improve without knowing that
something needs improving.  Math helpfully points out when we are
wrong, so we can improve.  This may sound like a motivational
platitude, but I assure you, that even as a working scientist, I am
wrong MOST of the time, and days when I find that I wasn't wrong are
usually days in which I wasn't working very hard.

\item \textbf{When you try to figure something out, it is often best
  to try an experiment first.}

Understanding the Monty Hall problem just by thinking about it doesn't
work for most people, and there is NOTHING WRONG WITH THAT.  Many
people think that if they don't ``get'' something quickly, it must be
too hard for them, or they take it as evidence that this is not their
calling.  

Math is hard for everybody (some of us have just forgotten
about all the work we once did), and it's silly to think that you can
``think your way through'' everything.  Doing experiments, drawing
pictures, finding ways to visualize and think about a problem -- all
of these are essential parts of actually figuring stuff out.  

This book will help you learn more tools to understand and figure out
problems, sometimes using techniques that may seem like tedious
distractions.  You will find yourself wondering, ``Why won't they just
teach me the \emph{right} way to do it and be done with it, instead of
forcing us to do all this exploration?''

The sad truth is that as professors, we are no more able to anticipate
the kinds of problems you will actually face than we are to predict
the future.  I couldn't possibly teach you every single algorithm, so
you could follow it like a cookbook, even if I wanted to.  If I did,
where would you be when you were faced with your first genuine
problem, with no idea how to begin?  And, frankly, if I could put it
in a cookbook, then why wouldn't I just let a computer do it, rather
than try to teach fallible humans who will always be slower?
\end{itemize}










